Fast and robust ancestry inference using principal component analysis

### 2018-06-21 doc/paper/paper.tex

\newcommand{\bO}{\mathcal{O}}
\newcommand{\bR}{\mathbb{R}}
\newcommand{\ostu}{\operatorname{stu}}
\newcommand{\oref}{\operatorname{ref}}
\newcommand{\argmin}[1]{\underset{#1}{\operatorname{arg\;min}}\;}
\newcommand{\todo}[1]{{\color{red} (#1)}}

\title{Computational Methods for Principle Component Analysis Applied to
  Genotype Data}
\date{}

@@

To address these problems, we propose two alternative approaches, Adjusted Projection (AP) and Online ADP (OADP), implemented in the genoPCA software. Using random matrix theory results, AP asymptotically estimates and adjusts for the bias of SP. Moreover, OADP uses online singular value decomposition, which can greatly reduce the computation cost of ADP. We carried out extensive simulation studies to show that these alternative approaches are unbiased and that the computation times are 10-100 times faster than ADP when 2000 reference samples are used. We also applied our approaches to 500,000 study samples from the UK biobank data with 2,500 samples from the 1000 Genomes data as the reference. AP and OADP required 16 and 92 CPU hours, respectively, while the projected computation time of ADP is 1,933 CPU hours. Furthermore, when we only used the European reference samples in the 1000 Genomes to infer sub-European ancestry, SP clearly showed bias, unlike the proposed approaches. By using AP and OADP, we can infer ancestry and adjust for PS in GWAS robustly and efficiently

@@

ADP is regarded as the most accurate method for predicting study PC scores.
It is does not have SP's shrinkage bias.
However, the major disadvantage of ADP
is its high computational complexity.
In particular,
as the reference size increases,
the computational cost for a study sample
increases cubically.
Fortunately, this disadvantage can be mitigated by taking some computational shortcuts,
as in the OADP method.

@@

\begin{enumerate}
\item Compute $X^T X$.
  (Computational complexity: $\bO[n^2p]$.)  
\item Apply eigendecomposition to $X^T X$ to get $X^T X = V D^2 V^T$.
  ($\bO[n^3]$.)
\item Calculate $U_3 = X V_3 D_3^{-1}$,
  where $V_3$ is the first $k_3$ columns of $V$,
  and $D_3$ is the diagonal matrix of the first $k_3$ diagonal entries of $D$.
  ($\bO[k_3 n p]$.)
\item Calculate 
  \[
    L = U_3^T y \quad \text{and} \quad K = y^T H,
  \]
  where $H$ is the normalized  $y - U_3L$
  ($\bO[k_3 p]$.)
\item Calculate $Q^T Q$, where
  \[
    Q = 
    \begin{bmatrix}
      D_3 & L \\
      0 & K
    \end{bmatrix}.
  \]
  ($\bO[k_3^3]$.)
\item Apply eigendecomosition to $Q^T Q$ to get $Q^T Q = \ddot{V}_3 \ddot{D}^2_3 \ddot{V}^T_3$.
  ($\bO[k_3^3]$.)
\item Calculate
  \[
    \tilde{V}_3 =
    \begin{bmatrix}
      V_3 & 0 \\
      0 & 1
    \end{bmatrix}
    \ddot{V}_3.
  \]
  ($\bO[nk_3^2]$.)
\item Project the first $k_2$ columns of $\tilde{V}_3$ to the first $k_1$ columns of $V$ by using Procrustes analysis to get the transformed PC scores of the study individual,
  where $k_1 < k_2  < k_3 \ll n$.
  (Computational complexity is ignorable if $k_2$ is not large compared to $n$.)
  \item Go to step 4 for the next study individual.
\end{enumerate}

Total computational complexity: $\bO[n^2p + n^3 + k_3np + m(k_3p + k_3^3 + nk_3^2]) \approx \bO[n^2p + m(k_3p + k_3^2n)]$, provided $k_3 \ll n \ll p$.

The computational complexity of OADP for the study individuals increases linearly with respect to the reference sample size, which makes the computation time of OADP close to those of SP.
The closeness between the results given by OADP and ADP will be empirically shown later.

@@

For the comparison of computational cost,
we recorded each method's runtime for analyzing the study samples.
(The runtime for PCA on the reference samples
and for reading or writing files are not included.)
All the methods were run on a single CPU core.

@@

\section{Results}

@@

For the simulation study,
the PC scores predicted by AP, OADP, and ADP
are very close to each other
for every choice of reference sample size.
Compared to their results,
SP exhibits a strong shrinkage
when the reference sample size is 1000.
As we gradually increases the reference sample size to 3000
by an increment of 500,
the magnitude of the shrinkage decreases
but is still visible.
This is consistent with the theoretical result
that the shrinkage bias decreases
as the ratio of
the reference sample dimension to the reference sample size
decreases \citep{dey2016asymptotic}.

The accuracy of the four methods were compared
by using their ADP errors and center errors.
Compared to SP,
the other three methods' errors are very small
and close to each other.
OADP had smaller errors than AP
for both the ADP errors and the center errors.
Interestingly, the center errors of ADP were slightly higher
than that of OADP.
As the reference size increased,
the errors of all the four methods decreased.
SP exhibited the fastest decrease in both errors,
and the four methods' difference in accuracy
became smaller and smaller.

For computational cost,
we observed that the runtime for ADP increases faster than linearly
as the reference sample size increases.
This is consistent
with the theoretical computational complexity of ADP,
which expects the runtime to increase at the rate of $\bO[n^3]$.
When the reference sample size reached 3000,
ADP's runtime was 27 times of OADP's
and 225 times of SP's and AP's.
In comparison, the runtimes of SP, AP, and OADP
increased only slightly,
which is consistent with their $\bO[1]$ and $\bO[n]$ computational complexity when the data dimension and study sample size are fixed.
SP and AP were the most computational efficient methods.
Their runtimes were approximately only one-eighth of OADP's.

@@

When we used the 500 randomly selected global samples
in the UK Biobank data,
we found that the PC scores predicted by the four methods' are almost identical.
The separation of the ethnic groups was clear,
and the shrinkage bias of SP was not obvious,
which is consistent with the theoretical understanding
that the shrinkage bias decreases
as the signal-to-noise ratio increases \todo{Find citation}.

However, when we focused on the European (EUR) samples
among these 500 UK Biobank samples,
(of which there are 475 many,)
the results given by SP were clearly inaccurate.
The samples from British in England and Scotland (GBR)
and those from Utah residents with northern and western European ancestry	(CEU)
drifted very close to NULL.
This caused SP to incorrectly classified almost all of them as CEU,
since most GBR reference samples are farther away from NULL.
A few of the study samples predicted by ADP to be GBR or CEU
were even classified as the Iberian population in Spain (IBS),
since the IBS reference samples are close to NULL.
Moreover, the shrinkage for the IBS
and for the Toscani in Italia (TSI) populations
caused significant errors within these two groups.
The IBS reference samples
are approximately in the middle
between the TSI reference samples and NULL.
Thus most of the the study samples predicted by ADP to be TSI
were classified by SP as IBS.
The same patterns for the accuracy
of SP, AP, and OADP
discussed above
were observed
when we used all the 488,000 study samples in the UK Biobank dataset
and those predicted to be Europeans among them.

For AP and OADP, their predicted PC1 and PC2 scores
for the European study samples
were almost identical
to those done by ADP.
The three methods' classification
agreed on most of the samples,
except for within the cluster of CEU and GBR.
Since many residents with northern and western European ancestry living in Utah
could originally migrated from Britain,
these two groups' ancestral history could be very similar
and thus difficult if not impossible to be separated by PCA.
Hence any slightest change
in the PC scores for these two groups' study samples
could easily change the classification from one group to the other.
This is a problem inherent to the population structure
and is not caused by the limitations of our methods.

Just as in the simulation study,
SP's errors were the greatest.
Its ADP error
was twice of SP's and 9 times of OADP's for the global samples,
and 14 times of SP's and OADP's for the European samples.
For the center errors,
SP's error was still much greater than the other three methods.
For the European samples,
ADP's center error was the lowest,
followed by OADP and AP.
Interestingly, this order of the three methods
was reversed for the global samples.

For the computational cost,
ADP's runtime was 1.5 times of OADP's and 10 times of AP's and SP's
when only the European samples
of the 500 random global study samples
and the 498 European reference samples of the 1000 Genomes data
were used.
This difference increased to 18 times, 103 times, and 108 times
when we studied the all the 500 random global samples
with all the 2492 1000 Genomes reference samples.
As we included all the 488,000 study samples in the UK Biobank data,
SP, AP, and OADP only took 13, 16, and 90 CPU hours,
while ADP was estimated (from its 500-sample runtime)
to take 1828 CPU hours,
which makes ADP very computational costly to use in practice.

@@

For the future, we would like
to improve the parallelization of our software.
Currently it uses the poor-man’s parallelism,
that is,
dividing the study dataset into multiple smaller files on the hard drive,
applying the analysis to each of them,
and merge the results.
This parallelism saves both the memory usage and the runtime.
However, we would like to utilize
more sophisticated parallel computation techniques
to further improve the performance of our software.

@@

\begin{figure}
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.98\linewidth]{img/ggsim3000_100000_200_2_1_100_1_sturef_ggsim3000_100000_3000_2_1_100_0_sp}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.98\linewidth]{img/ggsim3000_100000_200_2_1_100_1_sturef_ggsim3000_100000_3000_2_1_100_0_sp}
\end{subfigure}
\hfill
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.98\linewidth]{img/ggsim3000_100000_200_2_1_100_1_sturef_ggsim3000_100000_3000_2_1_100_0_oadp}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.98\linewidth]{img/ggsim3000_100000_200_2_1_100_1_sturef_ggsim3000_100000_3000_2_1_100_0_adp}
\end{subfigure}
\caption{
  The PC scores predicted by each of the four methods
  as the reference size varies.
  Reference samples are in colors
  and study samples are in black.
  The stars mark the centers of mass (geographic centers) of each population for the reference and study samples.
  The number of variants is 100,000,
  and the study sample size is 200.
  Only the top 4 PCs are calculated.
  The data is simulated by the GGS software \citep{mathieson2012differential}.
}
\end{figure}

@@

\begin{table}[h]
  \centering
  \begin{tabular}{|l|l|l|l|l|l|}
    \hline
    Reference Size & 1000 & 1500 & 2000 & 2500 & 3000 \\
    \hline
    \multicolumn{6}{|c|}{Study Runtime} \\
    \hline
    SP         &  17  & 17   & 16   & 18   & 18   \\
    AP         &  18  & 16   & 18  & 16   & 17  \\
    OADP       &  145  & 146   & 147   & 148   & 149   \\
    ADP        &  379  & 809 & 1526 & 2556 & 4062 \\
    \hline
    \multicolumn{6}{|c|}{Error (ADP)} \\
    \hline
    SP         &  188  & 139   & 112   & 91   & 84   \\
    AP         &  31  & 27   & 12  & 11   & 14  \\
    OADP       &  21  & 16   & 12   & 9   & 10   \\
    ADP        &  0  & 0 & 0 & 0 & 0 \\
    \hline
    \multicolumn{6}{|c|}{Error (centers)} \\
    \hline
    SP         &  335 & 161  & 113 & 71  & 60  \\
    AP         &  31  & 19   & 17  & 25  & 15  \\
    OADP       &  12  & 17   & 11  & 20  & 12  \\
    ADP        &  20  & 14   & 14  & 23  & 14  \\
    \hline
  \end{tabular}
  \caption{
    The runtimes and errors of the four methods
    as the reference size increases.
    Runtimes are measured in seconds.
    Errors (ADP) are the Euclidean distances from
    the PC scores predicted by each method
    to those predicted by ADP.
    Errors (centers) are the Euclidean distances
    from the centers of mass of the reference samples
    to the centers of mass of the study samples.
    The number of variants is 100,000,
    and the study sample size is 200.
    Only the top 4 PCs are calculated.
    The data is simulated by the GGS software \citep{mathieson2012differential}.
    }
\end{table}

\begin{figure}[h]
  \centering
  \includegraphics[width=.98\linewidth]{img/ggsim_runtimes.png}
  \centering
  \includegraphics[width=.98\linewidth]{img/ggsim_errors_adp.png}
  \centering
  \includegraphics[width=.98\linewidth]{img/ggsim_errors_ctr.png}
  \caption{
    The runtimes and errors of the four methods
    as the reference size increases.
    Runtimes are measured in seconds.
    Errors from ADP are the Euclidean distance
    from each method's result to ADP's result.
    Errors between centers are the Euclidean distance
    from the centers of mass of the reference populations
    to the corresponding centers of mass of the study populations.
    The number of variants is 100,000,
    and the study sample size is 200.
    Only the top 4 PCs are calculated.
    The data is simulated by the GGS software \citep{mathieson2012differential}.
  }
\end{figure}

\begin{table}[H]
  \centering
  \begin{tabular}{|r|r|r|r|r|r|r|r|}
    \hline
    & \multicolumn{3}{|c|}{Runtime} & \multicolumn{2}{|c|}{Error (ADP)} & \multicolumn{2}{|c|}{Error (centers)} \\
    \hline
    Samples & All (hrs) & Global (sec) & European (sec) & Global & European & Global & European \\
    \hline
    Sample size & 488,000 & 500 & 475 & 500 & 475 & 500 & 475 \\
    \hline
    SP   & 13    &  62   & 48  & 92  & 450 & 3156 & 2304 \\ 
    AP   & 16    &  65   & 48  & 48  & 33  & 2486 & 706  \\
    OADP & 90    &  363  & 349 & 10  & 32  & 2575 & 665  \\
    ADP  & 1828*  &  6739 & 518 & 0   & 0   & 2578 & 649  \\
    \hline
  \end{tabular}
  \caption{
    The study runtimes and erros for the four methods
    for 1) all the samples and 2) 500 randomly selected study samples
    in the UK Biobank data.
    Errors (ADP) are the Euclidean distances from
    the PC scores predicted by each method
    to those predicted by ADP.
    Errors (centers) are the Euclidean distances
    from the centers of mass of the reference populations
    to the corresponding centers of mass of the study populations
    predicted by the $k$-nearest neighbor method.
    For the global study,
    2492 samples from the 1000 Genomes data are used as the reference.
    For the European study,
    only the 498 European samples among them are used.
    The number of variants
    in common between the reference and study samples is 145,282.
    Only the top 4 PCs are calculated.
    *Estimated from the runtime of ADP for the 500 global samples.
    }
\end{table}

\begin{figure}[p]
  \centering
  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_sp}
  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_ap}
  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_oadp}
  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_adp}
  \caption{PC scores of 500 randomly selected global samples in UK Biobank data (circles) predicted by the four methods with the global samples in the 1000 Genomes data (squares) as the reference.}
\end{figure}

\begin{figure}[p]
  \centering
  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_pred_EUR_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_EUR_sp}
  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_pred_EUR_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_EUR_ap}
  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_pred_EUR_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_EUR_oadp}
  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_pred_EUR_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_EUR_adp}
  \caption{PC scores of the European in the 500 randomly selected global samples UK Biobank data (circles) predicted by the four methods with the European samples in the 1000 Genomes data (squares) as the reference.}
\end{figure}

\begin{figure}[p]
  \centering
  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_nchunks100_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_sp.png}
  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_nchunks100_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_ap.png}
  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_nchunks100_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_oadp.png}
  \caption{All the 488366 samples in the UK Biobank data.}
\end{figure}

>>>>

\newcommand{\bO}{\mathcal{O}}
\newcommand{\bR}{\mathbb{R}}
\newcommand{\ostu}{\operatorname{stu}}
\newcommand{\oref}{\operatorname{ref}}
\newcommand{\argmin}[1]{\underset{#1}{\operatorname{arg\;min}}\;}
\newcommand{\todo}[1]{{\color{red} (TODO: #1)}}

\title{
  Fast and robust ancestry inference
  using principal component analysis
}
\author{Daiwei Zhang, Rounak Dey, Seunggeun Lee}
\date{}

@@

To address these problems, we propose two alternative approaches, Adjusted Projection (AP) and Online ADP (OADP), implemented in the genoPCA software. Using random matrix theory results, AP asymptotically estimates and adjusts for the bias of SP. Moreover, OADP uses online singular value decomposition, which can greatly reduce the computation cost of ADP. We carried out extensive simulation studies to show that these alternative approaches are unbiased and that the computation times are 10-100 times faster than ADP when 2000 reference samples are used. We also applied our approaches to 500,000 study samples from the UK biobank data with 2,500 samples from the 1000 Genomes data as the reference. AP and OADP required 16 and 90 CPU hours, respectively, while the projected computation time of ADP is 1,828 CPU hours. Furthermore, when we only used the European reference samples in the 1000 Genomes to infer sub-European ancestry, SP clearly showed bias, unlike the proposed approaches. By using AP and OADP, we can infer ancestry and adjust for PS in GWAS robustly and efficiently

@@

ADP is regarded as the most accurate method for predicting study PC scores \todo{Find citation}.
It does not have SP's shrinkage bias.
However, the major disadvantage of ADP
is its high computational complexity.
In particular,
as the reference size increases,
the computational cost for a study sample
increases cubically.
Fortunately, this disadvantage can be mitigated by taking some computational shortcuts,
as in the OADP method.

@@

\begin{enumerate}
\item Compute $X^T X$.
  (Computational complexity: $\bO[n^2p]$.)  
\item Apply eigendecomposition to $X^T X$ to get $X^T X = V D^2 V^T$,
  where $D$ and $V$ only need their first $k$ columns.
  ($\bO[n^3]$.)
\item Calculate $U = X V D^{-1}$ but for the first $k$ columns only.
  ($\bO[k n p]$.)
\item Calculate 
  \[
    L = U^T y \quad \text{and} \quad K = y^T H,
  \]
  where $H$ is the normalized  $y - UL$
  ($\bO[k p]$.)
\item Calculate $Q^T Q$, where
  \[
    Q = 
    \begin{bmatrix}
      D & L \\
      0 & K
    \end{bmatrix}.
  \]
  ($\bO[k^3]$.)
\item Apply eigendecomosition to $Q^T Q$ to get $Q^T Q = \ddot{V} \ddot{D}^2 \ddot{V}^T$.
  ($\bO[k^3]$.)
\item Calculate
  \[
    \tilde{V} =
    \begin{bmatrix}
      V & 0 \\
      0 & 1
    \end{bmatrix}
    \ddot{V}.
  \]
  ($\bO[nk^2]$.)
\item Transform the first $k_2$ columns of $\tilde{V}$ to the first $k_1$ columns of $V$ by using Procrustes analysis to get the transformed PC scores of the study individual,
  where $k_1 < k_2  < k \ll n$.
  (Computational complexity is negligible if $k_2$ is small compared to $n$.)
  \item Go to step 4 for the next study individual.
\end{enumerate}

The total computational complexity is $\bO[n^2p + m(kp + k^2n)]$
provided $k \ll n \ll p$.
The computational complexity of OADP for the study individuals increases linearly with respect to the reference sample size,
which is much more efficient than ADP's cubic increasing rate.
The closeness between the results given by OADP and ADP
will be empirically shown in \Cref{sec:results}.

@@

For the comparison of computational cost,
we recorded each method's runtime for analyzing the study samples.
(The runtime for PCA on the reference samples
and for reading or writing files are not included.)
We used our genoPCA software,
which implemented SP, AP, and OADP in Python.
and TRACE by \citet{wang2015improved},
which implemented ADP in C++.
All the programs were run on a single CPU core.

@@

\section{Results}\label{sec:results}.

@@

For the simulation study,
the PC scores predicted by AP, OADP, and ADP
are very close to each other
for every choice of reference sample size (\Cref{fig:ggsim}).
Compared to their results,
SP exhibits a strong shrinkage
when the reference sample size is 1000.
As we gradually increases the reference sample size to 3000
by an increment of 500,
the magnitude of the shrinkage decreases
but is still visible.
This is consistent with the theoretical result
that the shrinkage bias decreases
as the ratio of
the reference sample dimension to the reference sample size
decreases \citep{dey2016asymptotic}.

The accuracy of the four methods were compared
by using their ADP errors and center errors (\Cref{tbl:ggsim}).
Compared to SP,
the other three methods' errors are very small
and close to each other.
OADP had smaller errors than AP
for both the ADP errors and the center errors.
Interestingly, the center errors of ADP were slightly higher
than that of OADP.
As the reference size increased,
the errors of all the four methods decreased.
SP exhibited the fastest decrease in both errors,
and the four methods' difference in accuracy
became smaller and smaller (\Cref{fig:ggsim_trend}).

For computational cost (\Cref{tbl:ggsim}),
we observed that the runtime for ADP increases faster than linearly
as the reference sample size increases.
This is consistent
with the theoretical computational complexity of ADP,
which expects the runtime to increase at the rate of $\bO[n^3]$ (\Cref{tbl:cplx}).
When the reference sample size reached 3000,
ADP's runtime was 27 times of OADP's
and 225 times of SP's and AP's.
In comparison, the runtimes of SP, AP, and OADP
increased only slightly,
which is consistent with their $\bO[1]$ and $\bO[n]$ computational complexity when the data dimension and study sample size are fixed.
SP and AP were the most computational efficient methods.
Their runtimes were approximately only one-eighth of OADP's.

@@

When we used the 500 randomly selected global samples
in the UK Biobank data,
we found that the PC scores predicted by the four methods' are almost identical (\Cref{fig:ukb_500}).
The separation of the ethnic groups was clear,
and the shrinkage bias of SP was not obvious,
which is consistent with the theoretical understanding
that the shrinkage bias decreases
as the signal-to-noise ratio increases \todo{Find citation}.

However, when we focused on the European (EUR) samples
among these 500 UK Biobank samples,
(of which there are 475 many,)
the results given by SP were clearly inaccurate (\Cref{fig:ukb_500_eur}).
The samples from British in England and Scotland (GBR)
and those from Utah residents with northern and western European ancestry	(CEU)
drifted very close to NULL.
This caused SP to incorrectly classified almost all of them as CEU,
since most GBR reference samples are farther away from NULL.
A few of the study samples predicted by ADP to be GBR or CEU
were even classified as the Iberian population in Spain (IBS),
since the IBS reference samples are close to NULL.
Moreover, the shrinkage for the IBS
and for the Toscani in Italia (TSI) populations
caused significant errors within these two groups.
The IBS reference samples
are approximately in the middle
between the TSI reference samples and NULL.
Thus most of the the study samples predicted by ADP to be TSI
were classified by SP as IBS.

For AP and OADP, their predicted PC1 and PC2 scores
for the European study samples
were almost identical
to those done by ADP.
The three methods' classification
agreed on most of the samples,
except for within the cluster of CEU and GBR.
Since many residents with northern and western European ancestry living in Utah
could originally migrated from Britain,
these two groups' ancestral history could be very similar
and thus difficult if not impossible to be separated by PCA.
Hence any slightest change
in the PC scores for these two groups' study samples
could easily change the classification from one group to the other.
This is a problem inherent to the population structure
and is not caused by the limitations of our methods.
The patterns discussed above
were still observed
when we used all the 488,000 study samples in the UK Biobank dataset
and those predicted to be Europeans among them (\Cref{fig:ukb,fig:ukb_eur}).

Just as in the simulation study,
SP's errors were the greatest (\Cref{tbl:ukb}).
Its ADP error
was twice of SP's and 9 times of OADP's for the global samples,
and 14 times of SP's and OADP's for the European samples.
For the center errors,
SP's error was still much greater than the other three methods.
For the European samples,
ADP's center error was the lowest,
followed by OADP and AP.
Interestingly, this order of the three methods
was reversed for the global samples.

For the computational cost,
ADP's runtime was 1.5 times of OADP's and 10 times of AP's and SP's
when only the European samples
of the 500 random global study samples
and the 498 European reference samples of the 1000 Genomes data
were used (\Cref{tbl:ukb}).
This difference increased to 18 times, 103 times, and 108 times
when we studied the all the 500 random global samples
with all the 2492 1000 Genomes reference samples.
As we included all the 488,000 study samples in the UK Biobank data,
SP, AP, and OADP only took 13, 16, and 90 CPU hours,
while ADP was estimated (from its 500-sample runtime)
to take 1828 CPU hours,
which makes ADP very computational costly to use in practice.

@@

From the simulation study and the study on the UK Biobank data,
we see that AP and OADP have achieved the accuracy of ADP
and the computational efficiency of SP.
The advantage of AP's and OADP's accuracy
is the most clear
when the ratio of the reference sample size
to the number of features is low
or the signal-to-noise ratio is low.
On the other hand,
the advantage of AP's and OADP's computational efficiency
is the most observable when the reference sample size is large.
After the comparison of the four methods,
we conclude that for a very small reference size,
SP is unreliable since the shrinkage would be great,
and ADP is a decent method to use
because its computational inefficiency is negligible,
but AP and OADP would still be much faster than ADP
and their accuracy would be very close to ADP's.
As the reference sample size increases
(with respect to the number of variants),
all the four methods' accuracy converge together,
and SP becomes more and more accurate,
although for genotype data,
due to its high dimensions,
the reference size must be extremely large
in order for SP to achieve an accuracy
that is at the same level as the other three methods.
Therefore, for the sample sizes
of the data sets that we have currently,
OADP and SP are the most robust and fast methods
for ancestry inference by PCA.

For the future, we would like
to improve the parallelism of our genoPCA software.
We are also interested in applying randomized SVD algorithms
such as the one by \citet{halko2011finding}
for PCA on the reference data,
since as the reference sample size becomes larger and larger,
it becomes impractical to load such a large matrix into memory
to compute its SVD or ED.

@@

\begin{figure}
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.98\linewidth]{img/ggsim3000_100000_200_2_1_100_1_sturef_ggsim3000_100000_3000_2_1_100_0_sp}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.98\linewidth]{img/ggsim3000_100000_200_2_1_100_1_sturef_ggsim3000_100000_3000_2_1_100_0_sp}
\end{subfigure}
\hfill
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.98\linewidth]{img/ggsim3000_100000_200_2_1_100_1_sturef_ggsim3000_100000_3000_2_1_100_0_oadp}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.98\linewidth]{img/ggsim3000_100000_200_2_1_100_1_sturef_ggsim3000_100000_3000_2_1_100_0_adp}
\end{subfigure}
\caption{
  The PC scores predicted by each of the four methods
  as the reference size varies.
  Reference samples are in colors
  and study samples are in black.
  The stars mark the centers of mass (geographic centers) of each population for the reference and study samples.
  The number of variants is 100,000,
  and the study sample size is 200.
  Only the top 4 PCs are calculated.
  The data is simulated by the GGS software \citep{mathieson2012differential}.
  \label{fig:ggsim}
}
\end{figure}

@@

\begin{table}[h]
  \centering
  \begin{tabular}{|l|l|l|l|l|l|}
    \hline
    Reference Size & 1000 & 1500 & 2000 & 2500 & 3000 \\
    \hline
    \multicolumn{6}{|c|}{Runtime (sec)} \\
    \hline
    SP         &  17  & 17   & 16   & 18   & 18   \\
    AP         &  18  & 16   & 18  & 16   & 17  \\
    OADP       &  145  & 146   & 147   & 148   & 149   \\
    ADP        &  379  & 809 & 1526 & 2556 & 4062 \\
    \hline
    \multicolumn{6}{|c|}{Error (ADP)} \\
    \hline
    SP         &  188  & 139   & 112   & 91   & 84   \\
    AP         &  31  & 27   & 12  & 11   & 14  \\
    OADP       &  21  & 16   & 12   & 9   & 10   \\
    ADP        &  0  & 0 & 0 & 0 & 0 \\
    \hline
    \multicolumn{6}{|c|}{Error (centers)} \\
    \hline
    SP         &  335 & 161  & 113 & 71  & 60  \\
    AP         &  31  & 19   & 17  & 25  & 15  \\
    OADP       &  12  & 17   & 11  & 20  & 12  \\
    ADP        &  20  & 14   & 14  & 23  & 14  \\
    \hline
  \end{tabular}
  \caption{
    The study runtimes and errors of the four methods
    as the reference size increases.
    Runtimes are measured in seconds.
    Errors (ADP) are the Euclidean distances from
    the PC scores predicted by each method
    to those predicted by ADP.
    Errors (centers) are the Euclidean distances
    from the centers of mass of the reference samples
    to the centers of mass of the study samples.
    The number of variants is 100,000,
    and the study sample size is 200.
    Only the top 4 PCs are calculated.
    The data is simulated by the GGS software \citep{mathieson2012differential}.
    }
    \label{tbl:ggsim}
\end{table}

\begin{figure}[h]
  \centering
  \includegraphics[width=.98\linewidth]{img/ggsim_runtimes.png}
  \centering
  \includegraphics[width=.98\linewidth]{img/ggsim_errors_adp.png}
  \centering
  \includegraphics[width=.98\linewidth]{img/ggsim_errors_ctr.png}
  \caption{
    The runtimes and errors of the four methods
    as the reference size increases.
    Runtimes are measured in seconds.
    Errors from ADP are the Euclidean distance
    from each method's result to ADP's result.
    Errors between centers are the Euclidean distance
    from the centers of mass of the reference populations
    to the corresponding centers of mass of the study populations.
    The number of variants is 100,000,
    and the study sample size is 200.
    Only the top 4 PCs are calculated.
    The data is simulated by the GGS software \citep{mathieson2012differential}.
    \label{fig:ggsim_trend}
  }
\end{figure}

\begin{table}[H]
  \centering
  \begin{tabular}{|r|r|r|r|r|r|r|r|}
    \hline
    & \multicolumn{3}{|c|}{Runtime} & \multicolumn{2}{|c|}{Error (ADP)} & \multicolumn{2}{|c|}{Error (centers)} \\
    \hline
    Samples & All (hrs) & Global (sec) & European (sec) & Global & European & Global & European \\
    \hline
    Sample size & 488,000 & 500 & 475 & 500 & 475 & 500 & 475 \\
    \hline
    SP   & 13    &  62   & 48  & 92  & 450 & 3156 & 2304 \\ 
    AP   & 16    &  65   & 48  & 48  & 33  & 2486 & 706  \\
    OADP & 90    &  363  & 349 & 10  & 32  & 2575 & 665  \\
    ADP  & 1828*  &  6739 & 518 & 0   & 0   & 2578 & 649  \\
    \hline
  \end{tabular}
  \caption{
    The study runtimes and erros for the four methods
    for 1) all the samples and 2) 500 randomly selected study samples
    in the UK Biobank data.
    Errors (ADP) are the Euclidean distances from
    the PC scores predicted by each method
    to those predicted by ADP.
    Errors (centers) are the Euclidean distances
    from the centers of mass of the reference populations
    to the corresponding centers of mass of the study populations
    predicted by the $k$-nearest neighbor method.
    For the global study,
    2492 samples from the 1000 Genomes data are used as the reference.
    For the European study,
    only the 498 European samples among them are used.
    The number of variants
    in common between the reference and study samples is 145,282.
    Only the top 4 PCs are calculated.
    *Estimated from the runtime of ADP for the 500 global samples.
    }
    \label{tbl:ukb}
\end{table}

\begin{figure}[p]
  \centering
  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_sp}
  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_ap}
  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_oadp}
  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_adp}
  \caption{PC scores of 500 randomly selected global samples in UK Biobank data (circles) predicted by the four methods with the global samples in the 1000 Genomes data (squares) as the reference.}
  \label{fig:ukb_500}
\end{figure}

\begin{figure}[p]
  \centering
  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_pred_EUR_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_EUR_sp}
  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_pred_EUR_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_EUR_ap}
  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_pred_EUR_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_EUR_oadp}
  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_5c_pred_EUR_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_EUR_adp}
  \caption{PC scores of the European in the 500 randomly selected global samples UK Biobank data (circles) predicted by the four methods with the European samples in the 1000 Genomes data (squares) as the reference.}
  \label{fig:ukb_500_eur}
\end{figure}

\begin{figure}[p]
  \centering
  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_nchunks100_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_sp.png}
  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_nchunks100_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_ap.png}
  \includegraphics[width=0.98\textwidth]{img/ukb_snpscap_kgn_bial_orphans_nchunks100_sturef_kgn_bial_orphans_snps_ukb_snpscap_ukb_oadp.png}
  \caption{All the 488366 samples in the UK Biobank data.}
  \label{fig:ukb}
\end{figure}
